A Prediction and Analysis of the interplay between Graduates and Salary Prospects in the STEM Field

### 2024-01-29 doc/DataLiteracyStudentProject/paper.tex
increment

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data. Combined with a
partial autocorrelation function (PACF) the optimal value for the lag
is determined, see \Cref{sec:methods:lag}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{fig:accuracy model}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude in \Cref{sec:methods:causality} that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria as descriped in \citep{pacf}. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get a 95\% Confidence Interval. The optimal value according to PACF is $3$.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.704\% & 0.871\% & 2.658\% \\
        \hline
        Prediciton 2 & 2.786\% & 0.301\% & 0.286\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $3$.}
    \label{tab:resid1}
\end{table}
The percentage differences between the true values and the predicted ones, especially for the students, are quite high, see \Cref{tab:resid1}.
To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$, which is the lag. The abbreviation AIC
stands for Akaike Information Criterion and it tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
Both BIC and AIC return $6$ as an optimal lag value for our data set. Since we are focusing on predicting students, low errors in that domain
are more important to us. Minimizing these errors within the range from $3$ to $6$ leads to an optimal lag value of $5$.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $5$.}
    \label{tab:resid2}
\end{table}

>>>>

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data.
After processing (\Cref{sec:methods:processing}) the data, the model is optimized by different methods, see \Cref{sec:methods:lag}.
This ensures that the relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{fig:accuracy model}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude in \Cref{sec:methods:causality} that salary expectations have a significant impact on the amount of STEM students.
Further, we can model different salary development scenarios, and see how the students would likely react (\Cref{sec:results}).

@@

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two established methods, the PACF and information-based criteria as described in \citep{pacf}. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get a 95\% Confidence Interval. The optimal value according to PACF is $3$.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.704\% & 0.871\% & 2.658\% \\
        \hline
        Prediciton 2 & 2.786\% & 0.301\% & 0.286\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $3$.}
    \label{tab:resid1}
\end{table}
The percentage differences between the true values and the predicted ones, especially for the students, are quite high, see \Cref{tab:resid1}.
To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$, which is the lag. The abbreviation AIC
stands for Akaike Information Criterion and it tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
Both BIC and AIC return $6$ as an optimal lag value for our data set. Since we are focusing on predicting students, low errors in that domain
are more important to us. Minimizing these errors within the range from $3$ to $6$ leads to an optimal lag value of $5$.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $5$.}
    \label{tab:resid2}
\end{table}
